# Getting Started with `dplyr`

This lesson and the next one will introduce you to some powerful tools for exploratory data analysis, using the [`gapminder` dataset](https://cran.r-project.org/web/packages/gapminder/index.html) as an example. Simple descriptive statistics can be extremely powerful. When you read a newspaper columnist arguing that $X$ is to blame for the massive change in $Y$, how often do you take a look at the data to see if $Y$ is *really changing at all*? Or if it's actually changing in the *opposite direction*? These are easy wins and we should make the most of them, particularly given that [many common misconceptions are *predictable*](https://upgrader.gapminder.org/). 

Today we'll focus on basic data manipulation and summary statistics using the [`dplyr` package](https://dplyr.tidyverse.org/), part of the [Tidyverse](https://www.tidyverse.org/) family of R packages. We'll make heavy use of `dplyr` and other Tidyverse packages throughout these lessons. This lesson will only scratch the surface of `dplyr`, although we'll also pick up a few more `dplyr` tricks in later lessons as well. A good resource for learning more is [R for Data Science](https://r4ds.had.co.nz/). The [dplyr cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf) is also extremely helpful. 

So what is `dplyr` and why should you use it? There are three popular toolkits for data manipulation in R. The first is so-called "base R." This simply means the built-in commands like `subset()` and operators like `[]`, `$`, and so on. You need to know base R and you'll end up using it frequently, but it's simply not a great tool for complicated data manipulation. If you wrangle your data using base R commands, your code will be complicated, hard to read, and hard to maintain. The two main alternatives are [`dplyr`](https://dplyr.tidyverse.org/) and [`data.table`](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html). As you might expect, [opinions differ](https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly) on which of these is the "better" choice. Having used and taught both packages, here's my summary of the pros and cons. The `data.table` package is *extremely fast*. If you routinely work with very large datasets (a million rows or more), it's worth learning `data.table`. But `data.table` syntax is a bit arcane, and can off-putting to newcomers. The key advantage of `dplyr` is that it's easy to learn and its syntax is *extremely intuitive and easy to read*. It also has many helpful features that `data.table` lacks. Given the importance of replicability and open science, this tips the balance in favor of `dplyr` for this book.


## Package Installation 
Before we can get started, you'll need to install two packages: `dplyr`, and `gapminder`. To do this, you can either click on the "Packages" tab in RStudio or use the command `install.packages()` at the R console, e.g. 
```{r,eval=FALSE}
install.packages('dplyr')
install.packages('gapminder')
```
You only need to install a package *once*, but you need to *load* it every time you want to use it.
To load a package, use the `library()` command, e.g.
```{r, message = FALSE}
library(dplyr)
library(gapminder)
```
Now we're ready to go! I could tell you all about the data contained in `gapminder`, but an important part of this book is helping you to become self-sufficient. So instead I will leave this as an exercise for the reader!

### Exercise
After loading `gapminder` enter the command `?gapminder` in the R console to view the R help file for this dataset. Read the documentation you find there to answer the following questions: 

1. How many rows and columns does `gapminder` contain?
2. What information is contained in each row and column?
3. What is the source of the data?


## What is a tibble?
The `dplyr` package uses a special operator called the *pipe*, written `%>%` to chain together commands called *verbs* that act on objects called *tibbles*. This probably sounds complicated, but it's easier that it looks. Over the next few sections we'll slowly unpack the preceding sentence by looking at a number of simple examples. But first things first: what is a tibble? Let's see what happens if we display the `gapminder` dataset:
```{r}
gapminder
```
If you're used to working with dataframes in R, this may surprise you. Rather than trying to print all `nrow(gapminder)` rows on the screen, R helpfully shows us a useful summary of the information contained in `gapminder`. This is because `gapminder` is *not* a dataframe; it's a *tibble*, often abbreviated *tbl*. For our purposes, all you really need to know about tibbles is that they are souped up versions of R dataframes that are designed to work seamlessly with `dplyr`. To learn more see the chapter "Tibbles" in [*R for Data Science*](https://r4ds.had.co.nz/tibbles.html).

## Filter Rows with `filter`
We're ready to learn our first `dplyr` verb: `filter` selects rows.
Here's an example:
```{r}
gapminder %>% 
  filter(year == 2007)
```
Compare the results of running this command to what we got when we typed `gapminder` into the console above.
Rather than displaying the whole dataset, now R is only showing us the 142 rows for which the column `year` has a value of `2007`.
So how does this work? The pipe operator `%>%` "pipes" the tibble `gapminder` into the function `filter()`. The argument `year == 2007` tells `filter()` that it should find all the rows such that the logical condition `year == 2007` is `TRUE`.

Oh no! Have we accidentally deleted all of the other rows of `gapminder`?
No: we haven't made any changes to `gapminder` at all.
If you don't believe me try entering `gapminder` at the console.
All that this command does is *display* a subset of `gapminder`.
If we wanted to store the result of running this command, we'd need to assign it to a variable, for example
```{r}
gapminder2007 <- gapminder %>% 
  filter(year == 2007)
gapminder2007
```
We can also use `filter` to subset on two or more variables.
For example, here we display data for the US in 2007:
```{r}
gapminder %>% 
  filter(year == 2007, country == 'United States')
```
Notice that I always put a linebreak after the pipe operator `%>%` in my code examples. This isn't required to make the code run, but it's a good habit. As we start to write longer and more complicated `dplyr` commands, linebreaks will make it much easier to understand how our code works.

### Exercises
1. What is the difference between `x = 3` and `x == 3` in R?
`r hide("Show Solution")`
The first assigns the value `3` to the variable `x`; the second tests whether `x` is equal to `3` and returns either `TRUE` or `FALSE`.
`r unhide()`
2. Write code that uses `filter` to choose the subset of `gapminder` for which `year` is 2002.
```{r webex.hide="Show Solution"}
gapminder %>% filter(year == 2002)
```
3. When I displayed data for the US in 2007, I put quotes around `United States` but not around `year`. Why?
`r hide("Show Solution")`
This is because `year` contains numeric data while `country` contains character data, aka string data.
`r unhide()`
4. If you instead try to choose the subset with `year` equal to 2005, something will go wrong. Try it. What happens and why? 
`r hide("Show Solution")`
If you go back to the help file for `gapminder` you'll see that it only contains data for every fifth year. The year 2005 isn't in our dataset so `dplyr` will display an empty tibble:
```{r}
gapminder %>% 
  filter(year == 2005)
```
`r unhide()`
5. Write code that stores the data for Asian countries in a tibble called `gapminder_asia`. Then display this tibble. 
```{r webex.hide="Show Solution"}
gapminder_asia <- gapminder %>% 
  filter(continent == 'Asia')
gapminder_asia
```
5. Which country had the higher life expectancy in 1977: Ireland or Brazil?
Which had the higher GDP per capita?
```{r webex.hide="Show Solution"}
gapminder %>% 
  filter(year == 1977, country %in% c('Ireland', 'Brazil'))
```

## Sort data with `arrange`
Suppose we wanted to sort `gapminder` by `gdpPercap`.
To do this we can use the `arrange` command along with the pipe `%>%` as follows:
```{r}
gapminder %>% 
  arrange(gdpPercap)
```
The logic is very similar to what we saw above for `filter`. We pipe the tibble `gapminder` into the function `arrange()`. The argument `gdpPercap` tells `arrange()` that we want to sort by GDP per capita. Note that by default `arrange()` sorts in *ascending order*. If we want to sort in *descending* order, we use the function `desc()` as follows:
```{r}
gapminder %>% arrange(desc(gdpPercap))
```

### Exercises 
1. What is the lowest life expectancy in the `gapminder` dataset? Which country and year does it correspond to?
```{r webex.hide="Show Solution"}
gapminder %>% 
  arrange(lifeExp)
```
2. What is the highest life expectancy in the `gapminder` dataset? Which country and year does it correspond to?
```{r webex.hide="Show Solution"}
gapminder %>% 
  arrange(desc(lifeExp))
```


## Choose columns with `select`
We use the `select` verb to choose a subset of columns.
For example, to display only `pop`, `country`, and `year`, we would write
```{r}
gapminder %>% 
  select(pop, country, year)
```
Now suppose that we wanted to select every column *except* `pop`. 
Here's one way to do it:
```{r}
gapminder %>% 
  select(country, continent, year, lifeExp, gdpPercap)
```
but that takes a lot of typing! If there were more than a handful of columns in our tibble it would be very difficult to *deselect* a column in this way.
Fortunately there's a shortcut: use the minus sign 
```{r}
gapminder %>% select(-pop)
```
Just as we could when *selecting*, we can *deselect* multiple columns by separating their names with a comma:
```{r}
gapminder %>% select(-pop, -year)
```

It's easy to mix up the `dplyr` verbs `select` and `filter`.
Here's a handy mnemonic: `filteR` filters Rows while `seleCt` selects Columns. 
Suppose we wanted to select only the column `pop` from `gapminder`.

### Exercise
1. Select only the columns `year`, `lifeExp`, and `country` in `gapminder`.
```{r webex.hide="Show Solution"}
gapminder %>% 
  select(year, lifeExp, country)
```
2. Select all the columns *except* `year`, `lifeExp`, and `country` in `gapminder`.
```{r webex.hide="Show Solution"}
gapminder %>% 
  select(-year, -lifeExp, -country)
```



## The `summarize` verb
Suppose we want to calculate the sample mean of the column `lifeExp` in `gapminder`.
We can do this using the `summarize` verb as follows:
```{r}
gapminder %>% summarize(mean_lifeExp = mean(lifeExp))
```
Note the syntax: within `summarize` we have an *assignment statement*.
In particular, we assign `mean(lifeExp)` to the variable `mean_lifeExp`.
The key thing to know about `summarize` is that it always returns *collapses* a tibble with many rows into a single row.
When we think about computing a sample mean, this makes sense: we want to summarize the column `lifeExp` as a single number.
It doesn't actually make much sense to compute the mean of `lifeExp` because this involves averaging over different countries *and* different years.
Instead let's compute the mean for a single year: 1952:
```{r}
gapminder %>% 
  filter(year == 1952) %>%
  summarize(mean_lifeExp = mean(lifeExp))
```

We can use summarize to compute multiple summary statistics for a single variable, the same summary statistic for multiple variables, or both:
```{r}
gapminder %>% 
  filter(year == 1952) %>%
  summarize(mean_lifeExp = mean(lifeExp),
            sd_lifeExp = sd(lifeExp), 
            mean_pop = mean(pop))
```
Note that if we *don't* explicitly use an assignment statement, R will make up names for us based on the commands that we used:
```{r}
gapminder %>% 
  filter(year == 1952) %>%
  summarize(mean(lifeExp), median(lifeExp), max(lifeExp))
```


### Exercise
1. Use `summarize` to compute the 75th percentile of life expectancy in 1977.
```{r webex.hide="Show Solution"}
gapminder %>% 
  filter(year == 1977) %>%
  summarize(quantile(lifeExp, 0.75))
```
2. Use `summarize` to compute the 75th percentile of life expectancy among Asian countries in 1977.
```{r webex.hide="Show Solution"}
gapminder %>% 
  filter(year == 1977, continent == 'Asia') %>%
  summarize(quantile(lifeExp, 0.75))
```


## The `group_by` verb
The true power of `summarize` is its ability to compute grouped summary statistics in combination with another `dplyr` verb: `group_by`.
In essence, `group_by` allows us to tell `dplyr` that we don't want to work with the whole dataset at once; rather we want to work with particular *subsets* or groups.
The basic idea is similar to what we've done using `filter` in the past.
For example, to calculate mean population (in millions) and mean life expectancy in the year 2007, we could use the following code:
```{r}
gapminder %>%
  filter(year == 2007) %>%
  summarize(meanPop = mean(pop) / 1000000, meanLifeExp = mean(lifeExp))
```
Using `group_by` we could do the same thing for *all* years in the dataset at once:
```{r}
gapminder %>%
  group_by(year) %>%
  summarize(meanPop = mean(pop) / 1000000, meanLifeExp = mean(lifeExp))
```
Notice what has changed in the second code block: we replaced `filter(year == 2007)` with `group_by(year)`.
This tells `dplyr` that, rather than simply restricting attention to data from 2007, we want to form *subsets* (groups) of the dataset that correspond to the values of the `year` variable.
Whatever comes after `group_by` will then be calculated for these subsets.

Here's another example.
Suppose we wanted to calculate mean life expectancy and total population in each *continent* during the year 2007.
To accomplish this, we can chain together the `filter`, `group_by` and `summarize` verbs as follows:
```{r}
gapminder %>% 
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(meanPop = mean(pop) / 1000000, meanLifeExp = mean(lifeExp))
```

We can also use `group_by` to subset over multiple variables at once.
For example, to calculate mean life expectancy and total population in each continent *separately* for every year, we can use the following code:
```{r}
gapminder %>% 
  group_by(year, continent) %>%
  summarize(meanPop = mean(pop) / 1000000, meanLifeExp = mean(lifeExp))
```

### Exercise
1. Why doesn't the following code work as expected?
```{r,eval = FALSE}
gapminder %>%
  summarize(meanLifeExp = mean(lifeExp)) %>%
  group_by(year)
```
`r hide("Show Solution")`
The steps are carried out in the wrong order: we need to form groups *first* and then calculate our desired summaries.
`r unhide()`
2. Calculate the median GDP per capita in each continent in 1977.
```{r webex.hide="Show Solution"}
gapminder %>% 
  filter(year == 1977) %>%
  group_by(continent) %>%
  summarize(medGDPc = median(gdpPercap))
```
3. Repeat 2. but sort your results in descending order.
```{r webex.hide="Show Solution"}
gapminder %>% 
  filter(year == 1977) %>%
  group_by(continent) %>%
  summarize(medGDPc = median(gdpPercap)) %>%
  arrange(desc(medGDPc))
```
4. Calculate the mean and standard deviation of life expectancy for separately for each continent in every year *after* 1977. Sort your results in ascending order by the standard deviation of life expectancy.
`r hide("Show Solution")`
```{r webex.hide="Show Solution"}
gapminder %>% 
  filter(year > 1977) %>%
  group_by(continent, year) %>%
  summarize(meanGDPc = mean(gdpPercap), sdGDPc = sd(gdpPercap)) %>%
  arrange(sdGDPc)
```


## Understanding the pipe: `%>%`
Let's revisit the pipe, `%>%`, that we've used in the code examples above.
I told you that the command `gapminder %>% filter(year == 2007)` "pipes" the tibble `gapminder` into the function `filter()`.
But what exactly does this mean?
Take a look at the R help file for the `dplyr` function `filter`.
We see that `filter()` takes something called `.data` as its first argument.
Moving on to the "Arguments" section of the help file, we see that `.data` is "A tbl" i.e. a tibble.
To better understand what this means, let's try using `filter` *without* the pipe:
```{r}
filter(gapminder, year == 2007, country == 'United States')
```
Notice that this gives us *exactly* the same result as
```{r}
gapminder %>% 
  filter(year == 2007, country == 'United States')
```
In other words __*The pipe is gives us an alternative way of supplying the first argument to a function*__.
Let's try this with a more familiar R function: `mean`.
The first argument of `mean` is a vector `x`. 
So let's try using the pipe to compute the mean of some data:
```{r}
x <- c(1, 5, 2, 7, 2)
x %>% mean
```
The pipe supplies a function with its *first* argument.
If we want to specify additional arguments, we need to do so within the function call itself.
For example, here's how we could use the pipe to compute the mean after dropping missing observations:
```{r}
y <- c(1, 5, NA, 7, 2)
y %>% 
  mean(na.rm = TRUE)
```
One important note about the pipe: it's *not* a base R command.
Instead it's a command provided by the package `Magrittr`.
(If you're familiar with the Belgian painter Magritte, you may realize that the name of this package is quite witty!)
This package is installed automatically along with `dplyr`.

### Exercise

1. Write R code that uses the pipe to calculate the sample variance of `z <- c(4, 1, 5, NA, 3)` excluding the missing observation from the calculation.
```{r webex.hide="Show Solution"}
z <- c(4, 1, 5, NA, 3)
z %>% var(na.rm = TRUE)
```
2. Write code to arrange `gapminder` by `lifeExp`, first in increasing order and then in decreasing order, *without* using the pipe.
```{r webex.hide="Show Solution"}
arrange(gapminder,lifeExp)
arrange(gapminder, desc(lifeExp))
```


## Chaining commands
In the examples we've looked at so far, the pipe doesn't seem all that useful: it's just an alternative way of specifying the first argument to a function.
The true power and convenience of the pipe only becomes apparent we need to *chain* a series of commands together.
For example, suppose we wanted to display the 1952 data from `gapminder` sorted by `gdpPercap` in descending order.
Using the pipe, this is easy:
```{r}
gapminder %>% 
  filter(year == 1952) %>% 
  arrange(desc(gdpPercap))
```
Notice how I split the commands across multiple lines.
This is good practice: it makes your code much easier to read.
So what's happening when we chain commands in this way? 
The first step in the chain `gapminder %>% filter(year == 1952)` returns a tibble: the subset of `gapminder` for which `year` is `1952`.
The next step `%>% arrange(gdpPercap)`  pipes this *new* tibble into the function `arrange()`, giving us the desired result.
I hope you agree with me that this is pretty intuitive: even if we didn't know anything about `dplyr` we could *almost* figure out what this code is supposed to do.
In stark contrast, let's look at the code we'd have to use if we wanted to accomplish the same task *without* using the pipe:
```{r}
arrange(filter(gapminder, year == 1952), 
        desc(gdpPercap))
```
There are may reasons why this code is harder to read, but the most important one is that the commands `arrange` and `filter` have to appear in the  code in the *opposite* of the order in which they are actually being carried out.
This is because parentheses are evaluated from *inside to outside*.
This is what's great about the pipe: it lets us write our code in a way that accords with the actual order of the steps we want to carry out.

### Exercise 

1. What was the most populous European country in 1992? Write appropriate `dplyr` code using the pipe to display the information you need to answer this question.
```{r webex.hide="Show Solution"}
gapminder %>%
  filter(year == 1992, continent == 'Europe') %>%
  arrange(desc(pop))
```
2. Re-write your code from part 1. *without* using the pipe. 
```{r webex.hide="Show Solution"}
arrange(filter(gapminder, year == 1992, continent == 'Europe'), desc(pop))
```

## Change an existing variable or create a new one with `mutate`
It's a little hard to read the column `pop` in `gapminder` since there are so many digits.
Suppose that, instead of raw population, we wanted to display population *in millions*.
This requires us to `pop` by `1000000`, which we can do using the function `mutate()` from `dplyr` as follows:
```{r}
gapminder %>% 
  mutate(pop = pop / 1000000)
```
Note the syntax here: within `mutate()` we have an assignment statement, namely `pop = pop / 1000000`. 
This tells R to calculate `pop / 1000000` and assign the result to `pop`, in place of the original variable.

We can also use `mutate()` to create a new variable.
The `gapminder` dataset doesn't contain overall GDP, only GDP per capita.
To calculate GDP, we need to multiply `gdpPercap` by `pop`.
But wait! 
Didn't we just change pop so it's expressed in millions?
No: we never *stored* the results of our previous command, we simply displayed them.
Just as I discussed above, unless you *overwrite* it, the original `gapminder` dataset will be unchanged.
With this in mind, we can create the `gdp` variable as follows:
```{r}
gapminder %>% 
  mutate(gdp = pop * gdpPercap)
```

### Exercise 

1. Explain why we used `=` rather than `==` in the `mutate()` examples above.   
`r hide("Show Solution")`
We used `=` because we're carrying out an *assignment operation*. In contrast `==` tests for equality, returning `TRUE` or `FALSE`.
`r unhide()`
2. Calculate life expectancy *in months* and use it to answer the following question: "which country in the Americas had the shortest life expectancy *in months* in the year 1962?" 
```{r webex.hide="Show Solution"}
gapminder %>%
  mutate(lifeExpMonths = 12 * lifeExp) %>%
  filter(year == 1962, continent == 'Americas') %>%
  arrange(lifeExpMonths)
```


