# How to Outsmart a Nobel Laureate 

## Is the Hot Hand an Illusion?

If you've read 2002 Nobel Laureate Daniel Kahneman's best-selling book [*Thinking Fast and Slow*](https://www.google.co.uk/books/edition/Thinking_Fast_and_Slow/AV9x8XakdV0C?hl=en&gbpv=1&dq=thinking+fast+and+slow&printsec=frontcover), you may remember this passage about the *hot hand illusion*, a supposed illustration of the human tendency to see patterns in random noise:

> Amos [Tversky] and his students Tom Gilovich and Robert Vallone caused a stir with their study of misperceptions of randomness in basketball. The "fact" that players occasionally acquire a hot hand is generally accepted by players, coaches, and fans. The inference is irresistible: a player sinks three or four baskets in a row and you cannot help forming the causal judgment that this player is now hot, with a temporarily increased propensity to score. Players on both teams adapt to this judgment—teammates are more likely to pass to the hot scorer and the defense is more likely to doubleteam. Analysis of thousands of sequences of shots led to a disappointing conclusion: there is no such thing as a hot hand in professional basketball, either in shooting from the field or scoring from the foul line. Of course, some players are more accurate than others, but the sequence of successes and missed shots satisfies all tests of randomness. The hot hand is entirely in the eye of the beholders, who are consistently too quick to perceive order and causality in randomness. The hot hand is a massive and widespread cognitive illusion.

The research that Kahneman mentions was published in a famous paper by [Gilovich, Vallone & Tversky (1985)](https://labs.la.utexas.edu/gilden/files/2016/04/Gilo.Vallone.Tversky.pdf), and later summarized for a general audience in [Gilovich & Tversky (1989)](http://www.medicine.mcgill.ca/epidemiology/hanley/c323/hothand.pdf). The abstract of the original paper says it all:

> Basketball players and fans alike tend to believe that a player's chance of hitting a shot are greater following a hit than following a miss on the previous shot. However, detailed analyses of the shooting records of the Philadelphia 76ers provided no evidence for a positive correlation between the outcomes of successive shots. The same conclusions emerged from free-throw records of the Boston Celtics, and from a controlled shooting experiment with the men and women of Cornell's varsity teams. The outcomes of previous shots influenced Cornell players' predictions but not their performance. The belief in the hot hand and the "detection" of streaks in random sequences is attributed to a general misconception of chance according to which even short random sequences are thought to be highly representative of their generating process.

Between 1985 and 2011, when Kahneman's book was published, this result was replicated numerous times under a variety of different conditions, making it one of the better-documented biases in human decision-making. But it turns out that the hot-hand illusion is *itself* an illusion.  

In a recent issue of *Econometrica*, [Miller \& Sanjurjo (2018)](https://onlinelibrary.wiley.com/doi/abs/10.3982/ECTA14943) point out a subtle but consequential error in the statistical tests used in the hot hand literature. It turns out that these tests were biased *against* detecting evidence of a hot hand, even if it did in fact exist. Correcting this mistake and re-analyzing Gilovich, Vallone and Tversky's original dataset "reveals significant evidence of streak shooting, with large effect sizes." The hot hand is real, and the literature has now shifted to trying to estimate the *size* of the effect.^[See for example [this write-up](https://www.nber.org/digest-202202/exploring-hot-hand-basketball) of [Lantis & Nessen (2021)](https://www.nber.org/papers/w29468) in the [February 2022 NBER digest](https://www.nber.org/digest-2022-02).]

There are some helpful lessons that we can draw from this episode. First, we all make mistakes--even Nobel Laureates! Second, successful replications tell us less than we might hope: they could easily *reproduce* the bias of the original study. But in my view, the real lesson of the hot hand affair is much simpler: **you should always run a simulation study.** 

The probabilistic error that led so many researchers to draw the wrong conclusion about the hot hand is really quite subtle.^[See [Miller \& Sanjurjo (2019)](https://pubs.aeaweb.org/doi/pdfplus/10.1257/jep.33.3.144) for a more accessible explanation that connects to several related probability puzzles.] But at the same time, anyone who knows basic programming could have detected the mistake in five minutes if only they had bothered to look. In economics and statistics, simulation is a *superpower*. It helps us to understand our models, check for mistakes, and make unexpected connections, some of which may even lead to new theoretical results. If you'll pardon my continuation of the Swiss Army Knife metaphor, simulation is the knife: arguably the most useful tool in your toolbox. In this lesson, we'll cover some basic tools for carrying out a simulation experiment in R and use them to shed some light on the illusion of the hot hand illusion.

## Drawing Random Data in R

Before we can use simulations to study the illusion of the hot hand illusion, we need to review the basics of drawing random data in R. We'll examine the functions `sample()`, `rbinom()` and `set.seed()` in detail. I'll also point you to a large number functions for simulating from well-known probability distributions in R. Finally, you'll have a chance to practice what you've learned by solving a few short exercises.

### `sample()`
R has many helpful built-in functions for making simulated random draws. The simplest is `sample()`, which makes `size` random draws without replacement from a vector `x`. To test this out, I'll create a very simple vector
```{r}
my_vector <- c('there', 'is', 'no', 'largest', 'integer')
```
The following line of code makes two draws *without replacement* from `my_vector`
```{r}
sample(x = my_vector, size = 2) # without replacement
```
If I run the same line of code again, I may not get the same result: it's random!^[Technically, "random" draws made on a computer are only [pseudorandom](https://en.wikipedia.org/wiki/Pseudorandomness). We'll discuss this further below.]
```{r}
sample(x = my_vector, size = 2) # without replacement
```

To draw *with replacement*, set `replace = TRUE`
```{r}
sample(x = my_vector, size = 7, replace = TRUE) # with replacement
```
As usual in R, the argument names `x`, `size`, and `replace` are optional. But it is considered good coding style to explicitly supply an argument name whenever we're *overriding* a function's default behavior. This makes it easier for anyone reading our code to understand what's happening. Since `sample()` defaults to making draws *without* replacement, it's a good idea to write `replace = TRUE` rather then simply `TRUE`. But even without writing `replace = `, the code will still work as long as we supply all of the arguments in the correct *order*:
```{r, error=TRUE}
sample(my_vector, 7, TRUE) # bad style
sample(my_vector, 7, replace = TRUE) # good style
```


### Probability Distributions in R 

As a programming language targeted at statistical applications, R supplies built-in functions for all of the most common probability distributions.^[For less common distributions, see [CRAN Task View: Probability Distributions](https://cran.r-project.org/web/views/Distributions.html)]
These functions follow a consistent naming convention. They being with either `d`, `p`, `q`, or `r` and are followed by an abbreviated name for a particular probability distribution. The prefix `d` denotes a *density* function (or mass function for a discrete distribution); `p` denotes a *cumulative distribution function* (CDF), `q` denotes a *quantile* function, and `r` denotes a function for making *random draws* from a particular distribution. For example: `dunif()` gives the probability density function of a uniform random variable, `pnorm()` gives the CDF of a normal random variable, `qchisq()` gives the quantile function of a Chi-squared, and `rbinom` allows us to make random draws from a Binomial distribution. The following table gives a full list of the relevant commands.

| R commands | Distribution |
|:-----------|:--------------|
| `d/p/q/rbeta` | [Beta](https://en.wikipedia.org/wiki/Beta_distribution) |
| `d/p/q/rbinom` | [Binomial](https://en.wikipedia.org/wiki/Binomial_distribution) |
| `d/p/q/rcauchy` | [Cauchy](https://en.wikipedia.org/wiki/Cauchy_distribution) | 
| `d/p/q/rchisq` | [Chi-Squared](https://en.wikipedia.org/wiki/Chi-squared_distribution) |
| `d/p/q/rexp` | [Exponential](https://en.wikipedia.org/wiki/Exponential_distribution) |
| `d/p/q/rf` | [F](https://en.wikipedia.org/wiki/F-distribution) |
| `d/p/q/rgamma` | [Gamma](https://en.wikipedia.org/wiki/Gamma_distribution) |
| `d/p/q/rgeom` | [Geometric](https://en.wikipedia.org/wiki/Geometric_distribution) |
| `d/q/p/rhyper` | [Hypergeometric](https://en.wikipedia.org/wiki/Geometric_distribution) |
| `d/p/q/rlogis` | [Logistic](https://en.wikipedia.org/wiki/Logistic_distribution) |
| `d/p/q/rlnorm` | [Log Normal](https://en.wikipedia.org/wiki/Lognormal_distribution) |
| `d/p/q/rnbinom` | [Negative Binomial](https://en.wikipedia.org/wiki/Negative_binomial_distribution) |
| `d/p/q/rnorm` | [Normal](https://en.wikipedia.org/wiki/Normal_distribution) |
| `d/p/q/rpois` | [Poisson](https://en.wikipedia.org/wiki/Poisson_distribution) |
| `d/p/q/rt` | [Student's t](https://en.wikipedia.org/wiki/Student%27s_t-distribution) |
| `d/p/q/runif` | [Uniform](https://en.wikipedia.org/wiki/Continuous_uniform_distribution) |
| `d/p/q/rweibull` | [Weibull](https://en.wikipedia.org/wiki/Weibull_distribution) |

There's a single help file for all of the `d/p/q/r` functions for a particular distribution. For example, if you enter `?dbeta` at the console you'll be shown the help files for `dbeta()`, `pbeta()`, `qbeta()`, and `rbeta()`.

To get a feel for how these functions work, let's take a look at `rbinom()`, the function for drawing from a [Binomial distribution](https://en.wikipedia.org/wiki/Binomial_distribution). Recall that a Binomial$(m,p)$ random variable equals the number of heads in $m$ independent tosses of a coin with $\mathbb{P}(\text{Heads})=p$. Or to use a bit of probability jargon, it equals the number of *successes* in $m$ independent *Bernoulli trials*, each with probability of success $p$.^[A Bernoulli trial is a model for a possibly biased coin flip: if $X \sim \text{Bernoulli}(p)$ then $\mathbb{P}(X=1) = p$ and $\mathbb{P}(X=0) = 1-p$.] If $X$ is a Binomial random variable with parameters $m$ and $p$, traditionally written as $X \sim \text{Binomial}(m, p)$ then $X$ must take on a value in the set $\{0, 1, 2, ..., m\}$ and the probability that it takes on a particular value $x$ in this set is 
$$
\mathbb{P}(X = x) = \binom{m}{x} p^x (1 - p)^x
$$
The function `rbinom()` makes random draws with the probabilities given by this formula. Its takes three arguments: `size` is the number of trials, $m$ in the formula, `prob` is the probability of success, $p$ in the formula, and `n` is the desired number of Binomial draws. For example, we can make a single draw from a Binomial$(m = 10, p =1 /2)$ distribution as follows
```{r}
rbinom(n = 1, size = 10, prob = 0.5)
```
and fifteen draws from the same distribution by changing `n` to `10`
```{r}
rbinom(n = 15, size = 10, prob = 0.5)
```
It's important not to confuse `n` with `size`. The former tells R how many *Binomial* draws to make. The latter tells R the value of the parameter $m$ of the Binomial$(m, p)$ distribution.

Perhaps you remember that if $X \sim \text{Binomial}(m, p)$ then $\mathbb{E}[X] = mp$ and $\text{Var}= np(1-p)$. We can approximate these results numerically by simulating a large number of draws, say 5000, from a Binomial distribution:
```{r}
m <- 20
p <- 0.25
n_sims <- 5000
sim_draws <- rbinom(n_sims, m, p)
```
and then comparing the theoretical value for $\mathbb{E}(X)$ to a simulation-based approximation:
```{r}
c(EV_Theoretical = m * p, 
  EV_Simulation = mean(sim_draws))
```
and similarly for $\text{Var}(X)$
```{r}
c(Var_Theoretical = m * p * (1 - p), 
  Var_Simulation = var(sim_draws))
```
Reassuringly, our simulation results are very close to the theoretical values. They would be even closer if we used a larger value for `n_sims`.

### `set.seed()`

A key theme of this book is the importance of *reproducible research*. Anyone else who wants to check your work should be able to obtain *exactly the same results as you did* by running your code. But this seems to be at odds with the idea of simulating random data. For example, if I run `rbinom(10, 4, 0.6)` repeatedly, I'll most likely get different results each time:
```{r}
rbinom(10, 4, 0.6)
rbinom(10, 4, 0.6)
rbinom(10, 4, 0.6)
```
The function `set.seed()` allows us to ensure that we obtain the *same* simulation draws whenever we re-run the same simulation code. To use it, we simply choose a *seed*, any integer between negative and positive $2^{31} - 1$, and supply it as an argument to `set.seed()`. Simulation draws made on a computer aren't really random: they're only *pseudo-random*. This means that they "look" random and pass statistical tests for randomness in spite of being generated by a completely deterministic algorithm. Setting the *seed* sets the *initial condition* of the pseudorandom number generator. Because the algorithm is deterministic, the same initial condition always leads to the same sequence of draws. This is what allows us to replicate our simulation draws. Each time we make another draw, the seed changes. But we can always return it to its previous state using `set.seed()`. For example, suppose I set the seed to `1`. and re-run my code from above as follows
```{r}
set.seed(1)
x1 <- rbinom(10, 4, 0.6)
x1
```
If I run `rbinom(10, 4, 0.6)` again, I will most likely *not* get the same result, because the state of the pseudorandom number generator has changed:
```{r}
x2 <- rbinom(10, 4, 0.6)
x2
identical(x1, x2) # safe/reliable way to test if two objects are exactly equal
```
but if I reset the seed to `1` I'll obtain exactly the same result as before:
```{r}
set.seed(1)
x3 <- rbinom(10, 4, 0.6)
x3
identical(x3, x1)
```

*Whenever* you write simulation code, start by choosing a seed and adding the line `set.seed(MY-SEED-GOES HERE)` to the top of your R script. You'll often see people use `set.seed(12345)` or `set.seed(54321)`. When I'm not feeling lazy, I like to generate a *truly random number* to use as my seed. The website [random.org](https://random.org) provides free access to *bona fide* random numbers generated from atmospheric noise. The "True Random Number Generator" on the top right of their main page allows you to make uniform integer draws on a range from "Min" to "Max." Using the widest possible range, $\pm1\times 10^9$, I generated the seed `420508570` which I'll use in the following exercises.

### Exercises
Set your seed to `420508570` at the start of your solution code for each of these exercises to obtain results that match the solutions.

1. Run `sample(x = my_vector, size = 10)`. What happens and why?
`r hide("Show Solution")`
R will throw an error. You can't make ten draws *without replacement* from a set of five objects:
```{r, error=TRUE}
set.seed(420508570)
sample(x = my_vector, size = 10)
```
`r unhide()`
2. Write a line of code that makes five draws without replacement from the set of numbers $\{1, 2, 3, ..., 100\}$.
```{r webex.hide="Show Solution"}
set.seed(420508570)
sample(1:100, 5)
```
3. Create a vector of thirty elements called `urn` that represents an urn containing ten blue balls and twenty red balls. Draw five balls with replacement from `urn` and store the draws in a vector called `draws`. Then write a line of code to count up the number of blue balls in `draws`.
`r hide("Show Hint")`
Use `rep()` and `c()` to construct `urn`. Use `==` and `sum()` to count up the number of blue balls in `draws`. See the relevant help files for details, e.g. `?rep`.
`r unhide()`
```{r webex.hide="Show Solution"}
set.seed(420508570)
urn <- c(rep('blue', 10), rep('red', 20))
draws <- sample(urn, 5, replace = TRUE)
draws
sum(draws == 'blue')
```
4. Make 1000 draws from a normal distribution with mean 5 and variance 4 and store them in a vector called `normal_sims`. Calculate the mean and variance of your draws, and plot a histogram.
`r hide("Show Hint")`
Consult the help file for `rnorm()` paying close attention to the fact that R specifies the normal distribution in terms of a mean and *standard deviation* rather than a mean and *variance*. You can plot a histogram with any number of functions: e.g. the base R function `hist()` or `qplot()` from the `ggplot2` package.
`r unhide()`
```{r webex.hide="Show Solution"}
set.seed(420508570)
normal_sims <- rnorm(1000, 5, 2) # Variance = 4; Standard Dev. = 2
mean(normal_sims)
var(normal_sims)
ggplot2::qplot(normal_sims, bins = 25)
```
5. There is no built-in R function called `rbern()` for simulating draws from the [Bernoulli Distribution](https://en.wikipedia.org/wiki/Bernoulli_distribution) with probability of success $p$. Write one of your own and use it to make ten Bernoulli(0.8) draws. Your function `rbern()` should take two arguments: the number of draws `n` and the probability of success `p`.
`r hide("Show Hint")`
There are various ways to do this. The simplest is by setting the arguments of `rbinom()` appropriately.
`r unhide()`
```{r webex.hide="Show Solution"}
rbern <- function(n, p) {
# Make n random draws from a Bernoulli(p) distribution
  rbinom(n, size = 1, prob = p)
}
set.seed(420508570)
rbern(10, 0.8)
```

## The Skeleton of a Simulation Study

While the specific details will vary, nearly every simulation study has the same basic structure: 

1. Generate simulated data.
2. Calculate an estimate from the simulated data.
3. Repeat steps 1 and 2 many times, saving each of the estimates.
4. Summarize the results.

Thinking in terms of this structure helps us to write code that is easier to understand, easier to generalize, and faster to run. The key is to break these steps down into *functions* that carry out a single, well-defined task. Generally these will include:

a. A function to generate simulated data.
b. A function to calculate an estimate from the data.
c. A function that repeatedly calls a. and b. and summarizes the results.

This may sound a bit abstract, so let's see how it works in practice. My introductory statistics students often ask me why the sample variance, $S^2$, divides by $(n-1)$ rather than the sample size $n$:
$$
S^2 \equiv \frac{1}{n-1}\sum_{i=1}^n (X_i - \bar{X}_n)^2
$$
The answer is that dividing by $(n-1)$ yields an *unbiased estimator*: if $X_1, ..., X_n$ are a random sample from a population with mean $\mu$ and variance $\sigma^2$, then $\mathbb{E}[S^2] = \sigma^2$. So what would happen if we divided by $n$ instead? Consider the estimator $\widehat{\sigma}^2$ defined by
$$
\widehat{\sigma}^2 \equiv \frac{1}{n}\sum_{i=1}^n (X_i - \bar{X}_n)^2.
$$
If $X_i \sim \text{Normal}(\mu, \sigma^2)$ then $\widehat{\sigma}^2$ is in fact the *maximum likelihood estimator* for $\sigma^2$. With a bit of algebra, we can show that $\mathbb{E}[\widehat{\sigma}^2] = (n-1)\sigma^2/n$.^[To see this, first rewrite $\sum_{i=1}^n (X_i - \bar{X}_n)^2$ as $\sum_{i=1}^n (X_i - \mu)^2 - n(\bar{X}_n - \mu)^2$. This step is just algebra. Then take expectations, using the fact that the $X_i$ are independent and identically distributed.]
It follows that
$$
\text{Bias}(\widehat{\sigma}^2) \equiv \mathbb{E}[\widehat{\sigma}^2 - \sigma^2] = -\sigma^2/n
$$
so $\widehat{\sigma}^2$ is biased *downwards*. Because the bias goes to zero as the sample size grows, however, it is still a consistent estimator of $\sigma^2$.

Another way to see that $\widehat{\sigma}^2$ is biased is by carrying out a simulation study. To do this, we generate data from a distribution with a *known* variance and calculate $\widehat{\sigma}^2$. Then we generate a *new* dataset from the same distribution and again calculate the corresponding value of $\widehat{\sigma}^2$. Repeating this a large number of times, we end up with many estimates $\widehat{\sigma}^2$, each based on a dataset of the same size drawn independently from the same population. This collection of estimates gives us an *approximation* to the sampling distribution of $\widehat{\sigma}^2$. Using this approximation, we can get a good estimate of $\text{Bias}(\widehat{\sigma}^2)$ by comparing the sample mean of our simulated estimates $\widehat{\sigma}^2$ to the *true* variance $\sigma^2$. 

The first thing we need is a function to generate simulated data. Let's draw the $X_1, ..., X_n$ from a normal distribution with mean zero and variance `s_sq`. To do this, we write a simple R function as follows:
```{r}
draw_sim_data <- function(n, s_sq) {
  rnorm(n, sd = sqrt(s_sq))
}
```
The nice thing about writing such a function is that we can test that it's working correctly. For example, suppose you were worried that `draw_sim_data` does *not* in fact generate `n` draws from a normal distribution with mean zero and variance `s_sq`. Then you could simply draw a large sample and check! Everything seems to work as expected:
```{r}
set.seed(420508570)
test_sims <- draw_sim_data(5000, 9)
length(test_sims)
mean(test_sims)
var(test_sims)
qqnorm(test_sims)
qqline(test_sims)
```




```{r}
get_estimate <- function(x) {
  sum((x - mean(x))^2) / length(x) # divides by n not (n-1)
}
get_bias <- function(n, s_sq, n_reps = 5000) {
  draw_sim_replication <- function() {
    sim_data <- draw_sim_data(n, s_sq)
    get_estimate(sim_data)
  }
  sim_replications <- replicate(n_reps, draw_sim_replication())
  mean(sim_replications - s_sq)
}
set.seed(420508570)
get_bias(5, 1)
get_bias(10, 2)
sim_parameters <- expand.grid(n = 3:10, s_sq = c(1, 2, 3))
sim_results <- lapply(1:nrow(sim_parameters), 
                      function(i) get_bias(n = sim_parameters$n[i],
                                           s_sq = sim_parameters$s_sq[i]))
sim_results <- do.call(rbind, sim_results)
cbind(sim_parameters, sim_results)
```

## Projects

### The Hot Hand Revisited
In the lesson we generated iid data. Now try simulating from a first-order Markov chain for various different parameters. How do the results change?

### Which estimator of the variance is *really* better?
Compare the RMSE of two variance estimators: the unbiased estimator and the MLE.

### Optimal Stopping
Simulate the secretary problem; try to find the optimal stopping rule.

## DRAFT MATERIAL

**Should I introduce `knitr` and `Rmarkdown` in this lesson or the next one?**

1. Some basic simulation commands in R: `sample()`, `rbinom()`, `rnorm()`, etc. Read the help files.
2. Write a function `draw_sim_data()` that makes 100 Bernoulli(1/2) draws. Optional arguments `p` and `n`? This simulates data when there is *no hot hand*
3. `set.seed()` what does it do?
4. Think about how to calculate the estimator: fraction of times that three ones are followed by another one compared to another zero. Suppose you had a function `is_after_3_ones()` that took a vector of 0 and 1 and returned, for each element, whether it is after three ones. How would you use it? Write this function.
5. Put everything together with `replicate()` to do a simple sim for $p=1/2$ and $n = 100$.
6. How about trying different values of $n$ and $p$? Need to keep results organized: `apply()` family of functions (or maybe the tidy equivalents?)
7. Try doing it in parallel with `mclapply()`. First explain the basic idea of parallel and why this is "embarrassingly parallel." Show them how to time the code, illustrate with `sys.sleep()`. 
8. For the students who finish very quickly, have some extensions: a markov chain DGP, and `is_after_k_ones()`

```{r}
dgp <- function(n = 100, p = 0.5) {
  rbinom(n, 1, p)
}

# Maybe have a challenge to write the version for after k ones, but start by
# asking them to do the one for after 3 ones
is_after_k_ones <- function(x, k) {
  out <- rep(NA)
  for(i in (k+1):length(x)) {
    out[i] <- sum(x[(i - k):(i - 1)]) == k
  }
  return(out)
}

get_est <- function(x) {
  #ones <- which(x == 1)
  #mean(x[ones + 1], na.rm = TRUE)
  mean(x[is_after_k_ones(x, 3)], na.rm = TRUE)
}

n_reps <- 1000
set.seed(1234)
sim_results <- replicate(n_reps, get_est(dgp()))
library(ggplot2)
qplot(sim_results)
mean(sim_results, na.rm = TRUE)

#ones <- which(x == 1)
#x[which(x == 1) + 1]
#runs <- rle(x)
#foo <- rle(x)
#str(foo)
#x
#foo$lengths
#foo$values
```


