# Mind the Gap Part II

Some kind of intro? Or should this be a separate lesson? 
**I think I should split this off into a separate lesson. Add some extra `dplyr` stuff above, taken from the Econ 224 problem sets.**

## A simple scatterplot using `ggplot2`
Now that we know the basics of `dplyr`, we'll turn our attention to graphics.
R has many powerful build-in graphics functions that may be familiar to you from Econ 103.
In this class, however, we'll use a very powerful package for statistical visualization called `ggplot2`.
There's nothing more for you to instead or load, since `ggplot2` is included in the `tidyverse` package, which you've already installed and loaded.
For more details on `ggplot2` see the chapter entitled "Data Visualisation" in *R for Data Science*.

We'll start off by constructing a subset of the `gapminder` dataset that contains information from the year 2007 that we'll use for our plots below.
```{r}
gapminder_2007 <- gapminder %>% 
  filter(year == 2007)
```

It takes some time to grow accustomed to `ggplot2` syntax, so rather than giving you a lot of detail, we're going to look at a series of increasingly more complicated examples.
Our first example will be a simple scatterplot using `gapminder_2007`. 
Each point will correspond to a single country in 2007.
Its x-coordinate will be GDP per capita and its y-coordinate will be life expectancy.
Here's the code:
```{r}
ggplot(gapminder_2007) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp))
```
We see that GDP per capita is a very strong predictor of life expectancy, although the relationship is non-linear. Notice how I add a linebreak after the `+`. This is analogous to how I always add a linebreak after the pipe `%>%`. While it isn't necessary for the code to run correctly, it improves readability.

### Exercise
1. Using my code example as a template, make a scatterplot with `pop` on the x-axis and `lifeExp` on the y-axis using `gapminder_2007`. Does there appear to be a relationship between population and life expectancy?
`r hide("Show Solution")`
There is no clear relationship between population and life expectancy based on the 2007 data:
```{r}
ggplot(gapminder_2007) + 
  geom_point(mapping = aes(x = pop, y = lifeExp))
```
`r unhide()`
2. Repeat 1. with `gdpPercap` on the y-axis.
`r hide("Show Solution")`
There is no clear relationship between population and GDP per capita based on the 2007 data:
```{r}
ggplot(gapminder_2007) + 
  geom_point(mapping = aes(x = pop, y = gdpPercap))
```
`r unhide()`


## Plotting on the log scale
It's fairly common to transform data onto a log scale before carrying out further analysis or plotting.
If you've taken Econ 104, you may already be familiar with log transformations.
If not, don't worry about it: we'll discuss them later in the course.
For now, we'll content ourselves with learning how to transform the axes  in a `ggplot` to the log base 10 scale.
To transform the x-axis, it's as easy as adding a `+ scale_x_log10()` to the end of our command from above:
```{r}
ggplot(data = gapminder_2007) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp)) + 
  scale_x_log10()
```
Notice how I split the code across multiple lines and ended each of the intermediate lines with the `+`.
This makes things much easier to read.

### Exercise 
1. Using my code example as a template, make a scatterplot with the log base 10 of `pop` on the x-axis and `lifeExp` on the y-axis using the `gapminder_2007` dataset.
```{r webex.hide="Show Solution"}
ggplot(data = gapminder_2007) + 
  geom_point(mapping = aes(x = pop, y = lifeExp)) + 
  scale_x_log10()
```
2. Suppose that rather than putting the x-axis on the log scale, we wanted to put the *y-axis* on the log scale. Figure out how to do this, either by clever guesswork or a google search, and then redo my example with `gdpPercap` and `lifeExp`  with `gdpPercap` in levels and `lifeExp` in logs.
```{r webex.hide="Show Solution"}
ggplot(data = gapminder_2007) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp)) + 
  scale_y_log10()
```
3. Repeat 2. but with *both* axes on the log scale.
```{r webex.hide="Show Solution"}
ggplot(data = gapminder_2007) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp)) + 
  scale_x_log10() + 
  scale_y_log10()
```


## The color and size aesthetics
It's time to start unraveling the somewhat mysterious-looking syntax of `ggplot`.
To make a graph using `ggplot` we use the following template:
```{r eval=FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
replacing `<DATA>`, `<GEOM_FUNCTION>`, and `<MAPPINGS>` to specify what we want to plot and how it should appear.
The first part is easy: we replace `<DATA>` with the dataset we want to plot, for example `gapminder_2007` in the example from above.
The second part is also fairly straightforward: we replace `<GEOM_FUNCTION>` with the name of a function that specifies the kind of plot we want to make.
So far we've only seen one example: `geom_point()` which tells `ggplot` that we want to make a scatterplot.
We'll see more examples in a future lab.
For now, I want to focus on the somewhat more complicated-looking `mapping = aes(<MAPPINGS>)`.

The abbreviation `aes` is short for *aesthetic* and the code `mapping = aes(<MAPPINGS>)` defines what is called an *aesthetic mapping*.
This is just a fancy way of saying that it tells R how we want our plot to look.
The information we need to put in place of `<MAPPINGS>` depends on what kind of plot we're making.
Thus far we've only examined `geom_point()` which produces a scatterplot. 
For this kind of plot, the minimum information we need to provide is the location of each point.
For example, in our example above we wrote `aes(x = gdpPercap, y = lifeExp)` to tell R that `gdpPercap` gives the x-axis location of each point, and `lifeExp` gives the y-axis location.

When making a scatterplot with `geom_point` we are not limited to specifying the x and y coordinates of each point; we can also specify the size and color of each point.
This gives us a useful way of displaying more than two variables in a two-dimensional plot.
We do this using `aes`.
For example, let's use the color of each point to indicate `continent`
```{r}
ggplot(data = gapminder_2007) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
  scale_x_log10()
```
Notice how `ggplot` automatically generates a helpful legend.
This plot makes it easy to see at a glance that the European countries in 2007 ten to have high GDP per capita and high life expectancy, while the African countries have the opposite.
We can also use the *size* of each point to encode information, e.g. population:
```{r}
ggplot(data = gapminder_2007) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  scale_x_log10()
```

### Exercise
1. Would it make sense to set `size = continent`? What about setting `col = pop`?
   Explain briefly.
`r hide("Show Solution")`
Neither of these makes sense since `continent` is categorical and `pop` is continuous: `color` is useful for categorical variables and `size` for continuous ones.
`r unhide()`
2. The following code is slightly different from what I've written above. What is different. Try running it. What happens? Explain briefly.
```{r,eval=FALSE}
ggplot(gapminder_2007) +
  geom_point(aes(x = gdpPercap, y = lifeExp)) + 
  scale_x_log10()  
```
`r hide("Show Solution")`
It still works! You don't have to explicitly write `data` or `mapping` when using `ggplot`. I only included these above for clarity. In the future I'll leave them out to make my code more succinct. 
```{r}
ggplot(gapminder_2007) +
  geom_point(aes(x = gdpPercap, y = lifeExp)) + 
  scale_x_log10()  
```
`r unhide()`
3. Create a tibble called `gapminder_1952` that contains data from `gapminder` from 1952.
```{r webex.hide="Show Solution"}
gapminder_1952 <- gapminder %>% 
  filter(year == 1952)
```
4. Use `gapminder_1952` from the previous part  to create a scatter plot with population on the x-axis, life expectancy on the y-axis, and continent represented by the color of the points. Plot population on the log scale (base 10).
```{r webex.hide="Show Solution"}
ggplot(gapminder_1952) + 
  geom_point(aes(x = pop, y = lifeExp, color = continent)) +
  scale_x_log10()
```
5. Suppose that instead of indicating continent using color, you wanted all the points in the plot from 3. to be blue. Consult the chapter "Visualising Data" from *R for Data Science* to find out how to do this.
`r hide("Show Solution")`
When you want color to be a variable from your dataset, put `color = <VARIABLE>` *inside* of `aes`; when you simply want to set the colors of all the points, put `color = '<COLOR>'` *outside* of `aes`, for example
```{r}
ggplot(gapminder_1952) + 
  geom_point(aes(x = pop, y = lifeExp), color = 'blue') +
  scale_x_log10()
```
`r unhide()`



## Faceting - Plotting multiple subsets at once
Let's pick up where we left off in lab \#1, with a plot of GDP per capita and life expectancy in 2007:
```{r}
gapminder_2007 <- gapminder %>%
  filter(year == 2007)
ggplot(gapminder_2007) +
  geom_point(aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  scale_x_log10()  
```

This is an easy way to make a plot for a single year.
But what if you wanted to make the same plot for *every year* in the `gapminder` dataset?
It would take a lot of copying-and-pasting of the preceding code chunk to accomplish this.
Fortunately there's a much easier way: *faceting*.
In `ggplot2` a *facet* is a subplot that corresponds to a subset of your dataset, for example the year 2007.
We'll now use faceting to reproduce the plot from above for all the years in `gapminder` simultaneously:
```{r,fig.width = 7, fig.height=6}
ggplot(gapminder) +
  geom_point(aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  scale_x_log10() +
  facet_wrap(~ year)
```
Note the syntax here: in a similar way to how we added `scale_x_log10()` to plot on the log scale, we add `facet_wrap(~ year)` to facet by `year`. 
The tilde `~` is important: this has to precede the variable by which you want to facet.

Now that we understand how to produce it, let's take a closer look at this plot.
Notice how this plot allows us to visualize five variables *simultaneously*.
By looking at how the plots change over time, we see a pattern of increasing GDP per capita and life expectancy throughout the world between 1952 and 2007.
Notice in particular the dramatic improvements in both variables in the Asian economies.

### Exercise

1. What would happen if I were to run the following code? Explain briefly.
```{r, eval=FALSE}
ggplot(gapminder_2007) +
  geom_point(aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  scale_x_log10() +
  facet_wrap(~ year)
```
`r hide("Show Solution")`
We'll only get one facet since the tibble `gapminder_2007` only has data for 2007: 
```{r}
ggplot(gapminder_2007) +
  geom_point(aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  scale_x_log10() +
  facet_wrap(~ year)
```
`r unhide()`
2. Make a scatterplot with data from `gapminder` for the year 1977.
   Your plot should be faceted by continent with GDP per capita on the log scale on the x-axis, life expectancy on the y-axis, and population indicated by the size of each point.
```{r webex.hide="Show Solution"}
gapminder_1977 <- gapminder %>% 
  filter(year == 1977)
ggplot(gapminder_1977) +
  geom_point(aes(x = gdpPercap, y = lifeExp, size = pop)) +
  scale_x_log10() +
  facet_wrap(~ continent)
```
3. What would happen if you tried to facet by `pop`? Explain briefly.
`r hide("Show Solution")`
You'll get something crazy if you try this. Population is continuous rather than categorical so every country has a different value for this variable. You'll end up with one plot for every country, containing a single point:
```{r,eval= FALSE}
# Not run: it takes a long time and looks nasty!
gapminder_1977 <- gapminder %>% 
  filter(year == 1977)
ggplot(gapminder_1977) +
  geom_point(aes(x = gdpPercap, y = lifeExp, color = continent)) +
  scale_x_log10() +
  facet_wrap(~ pop)
```
`r unhide()`



## Plotting summarized data
By combining `summarize` and `group_by` with `ggplot`, it's easy to make plots of grouped data.
For example, here's how we could plot total world population in millions from 1952 to 2007.
First we construct a tibble which I'll name `by_year` containing the desired summary statistic grouped by year and display it:
```{r}
by_year <- gapminder %>%
  mutate(popMil = pop / 1000000) %>%
  group_by(year) %>% 
  summarize(totalpopMil = sum(popMil))
by_year
```
Then we make a scatterplot using `ggplot`:
```{r}
ggplot(by_year) +
  geom_point(aes(x = year, y = totalpopMil)) 
```
Here's a more complicated example where we additionally use color to plot each continent separately:
```{r}
by_year_continent <- gapminder %>%
  mutate(popMil = pop / 1000000) %>%
  group_by(year, continent) %>%
  summarize(totalpopMil = sum(popMil))
by_year
ggplot(by_year_continent) +
  geom_point(aes(x = year, y = totalpopMil, color = continent)) 
```
Make sure you understand how the preceding example works before attempting the exercise.


### Exercise
1. What happens if you append `+ expand_limits(y = 0)` to the preceding `ggplot` code?  Why might this be helpful in some cases?
`r hide("Show Solution")`
The function `expand_limits()` lets us tweak the limits of our x or y-axis in a `ggplot`. In this particular example `expand_limits(y = 0)` ensures that the y-axis begins at zero. Without using this command, `ggplot` will choose the y-axis on its own so that there is no "empty space" in the plot. Sometimes we may want to override this behavior.
`r unhide()`
2. Make a scatter with average GDP per capita across all countries in `gapminder` in the y-axis and `year` on the x-axis.
```{r webex.hide="Show Solution"}
by_year <- gapminder %>%
  group_by(year) %>% 
  summarize(meanGDPc = mean(gdpPercap))
ggplot(by_year) +
  geom_point(aes(x = year, y = meanGDPc)) 
```
3. Repeat 2. broken down by continent, using color to distinguish the points. Put mean GDP per capita on the log scale.
```{r webex.hide="Show Solution"}
by_year <- gapminder %>%
  group_by(year, continent) %>% 
  summarize(meanGDPc = mean(gdpPercap))
ggplot(by_year) +
  geom_point(aes(x = year, y = meanGDPc, color = continent)) +
  scale_y_log10()
```


## Line plots
Thus far we've only learned how to make one kind of plot with `ggplot`: a scatterplot, which we constructed using `geom_scatter()`.
Sometimes we want to *connect* the dots in a scatterplot, for example when we're interested in visualizing a trend over time. 
The resulting plot is called a *line plot*.
To make one, simply replace `geom_scatter()` with `geom_line()`.
For example:
```{r}
by_year_continent <- gapminder %>%
  mutate(popMil = pop / 1000000) %>%
  group_by(year, continent) %>%
  summarize(totalpopMil = sum(popMil))
by_year
ggplot(by_year_continent) +
  geom_line(aes(x = year, y = totalpopMil, color = continent)) 
```

### Exercise 
Repeat exercise 5-3 with a line plot rather than a scatterplot.
```{r webex.hide="Show Solution"}
by_year <- gapminder %>%
  group_by(year, continent) %>% 
  summarize(meanGDPc = mean(gdpPercap))
ggplot(by_year) +
  geom_line(aes(x = year, y = meanGDPc, color = continent)) +
  scale_y_log10()
```

## Bar plots
To make a bar plot, we use `geom_col()`.
Note that the `x` argument of `aes` needs to be a *categorical variable* for a bar plot to make sense.
Here's a simple example:
```{r}
by_continent <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(meanLifeExp = mean(lifeExp))
ggplot(by_continent) +
  geom_col(aes(x = continent, y = meanLifeExp))
```
Sometimes we want to turn a bar plot, or some other kind of plot, on its side.
This can be particularly helpful if the x-axis labels are very long.
To do this, simply add `+ coord_flip()` to your `ggplot` command, for example:
```{r}
ggplot(by_continent) +
  geom_col(aes(x = continent, y = meanLifeExp)) +
  coord_flip()
```


### Exercise
Make a collection of bar plots faceted by year that compare mean GDP per capita across countries in a given year.
Orient your plots so it's easy to read the continent labels.
```{r webex.hide="Show Solution", fig.width = 7, fig.height = 6}
by_continent <- gapminder %>%
  group_by(continent) %>% 
  summarize(meanGDPc = mean(gdpPercap))
ggplot(by_year) +
  geom_col(aes(x = continent, y = meanGDPc)) +
  facet_wrap(~ year) +
  coord_flip()
```


## Histograms
To make a `ggplot2` histogram, we use the function `geom_histogram()`.
Recall from Econ 103 that a histogram summarizes a *single* variable at a time by forming non-overlapping bins of equal width and calculating the fraction of observations in each bin.
If we choose a different width for the bins, we'll get a different histogram.
Here's an example of two different bin widths:
```{r}
gapminder_2007 <- gapminder %>%
  filter(year == 2007)
ggplot(gapminder_2007) +
  geom_histogram(aes(x = lifeExp), binwidth = 5)
ggplot(gapminder_2007) +
  geom_histogram(aes(x = lifeExp), binwidth = 1)
```

### Exercise 

1. All of the examples we've seen that use `ggplot` *besides* histograms have involved specifying both `x` and `y` within `aes()`. Why are histograms different?
`r hide("Show Solution")`
This is because histograms only depict a single variable while the other plots we've made show two variables at once.
`r unhide()`
2. What happens if you don't specify a bin width in either of my two examples? 
`r hide("Show Solution")`
If you don't specify a bin width, `ggplot2` will pick one for you and possibly give you a warning suggesting that you pick a better bin width manually.
`r unhide()`
3. Make a histogram of GDP per capita in 1977. Play around with different bin widths until you find one that gives a good summary of the data.
`r hide("Show Solution")`
There's no obvious *right answer* for the bin width, but here's one possibility:
```{r}
gapminder1977 <- gapminder %>%
  filter(year == 1977)
ggplot(gapminder_1977) +
  geom_histogram(aes(x = gdpPercap), binwidth = 5000)
```
`r unhide()`
4. Repeat 3. but put GDP per capita on the log scale.
`r hide("Show Solution")`
You'll need a much smaller bin width when using the log scale, for example:
```{r}
ggplot(gapminder_1977) +
  geom_histogram(aes(x = gdpPercap), binwidth = 0.2) + 
  scale_x_log10()
```
`r unhide()`
5. Compare and contrast the two different histograms you've made.
`r hide("Show Solution")`
No right answer: it's a discussion question! But the idea is to see how taking logs gets rid of the huge positive skewness in GDP per capita.
`r unhide()`


## Boxplots 
The final kind of `ggplot` we'll learn how to produce is a boxplot, a visualization of the *five-number summary* of a variable: minimum, 25th percentile, median, 75th percentile, and maximum.
To make a boxplot in `ggplot` we use the function `geom_boxplot()`, for example:
```{r}
ggplot(gapminder_2007) +
  geom_boxplot(aes(x = continent, y = lifeExp))
```
Compared to histograms, boxplots provide less detail but allow us to easily compare across groups.

### Exercise 
1. What is the meaning of the little "dots" that appear in the boxplot above? Use a Google search to find out what they are and how they are computed.
`r hide("Show Solution")`
They are outliers: `ggplot` considers any observation that is more than 1.5 times the interquartile range away from the "box" to be an outlier, and adds a point to indicate it.
`r unhide()`
2. Use faceting to construct a collection of boxplots, each of which compares log GDP per capita across continents in a given year. Turn your boxplots sideways to make it easier to read the continent labels.
```{r, webex.hide="Show Solution", fig.width = 7, fig.height = 6}
ggplot(gapminder) +
  geom_boxplot(aes(x = continent, y = gdpPercap)) +
  facet_wrap(~ year) +
  scale_y_log10() +
  coord_flip() +
  ggtitle('GDP per Capita by Continent: 1952-2007')
```
3. Use a Google search to find out how to add a title to a `ggplot`. Use it to add a title to the plot you created in 2.
`r hide("Show Solution")`
Use `ggtitle('YOUR TITLE HERE')` as I did in my solution to 2. above. 
`r unhide()`



